The role of aneuploidy in the evolution of cancer drug resistance

### 2024-05-15 EvolutionaryRescue1.tex
Add files via upload

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: Yoav Ram \textcolor{green}{(e-mail: yoavram@tauex.tau.ac.il)}}

@@

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute \textcolor{green}{towards evolutionary rescue}.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} \textcolor{green}{and tested the predictions of their model with data from experiments with yeast and bacterium}.

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We \textcolor{green}{estimate the} effect of aneuploidy on the probability of evolutionary rescue when it acts as an \textcolor{green}{evolutionary stepping stone} between the \textcolor{green}{sensitive} and the \textcolor{green}{resistant} cancer cells. We also estimate the mean time \textcolor{green}{until the} tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of \textcolor{green}{pre-treatment} standing genetic variation on the \textcolor{green}{evolutionary} dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

The \textcolor{green}{cell population} at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the \textcolor{green}{population} (left column, see \Cref{figureAneuploidy}):

@@

We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division. 
\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division \textcolor{green}{(for $u\ll10^{-3}$ no adaptation happens and for $u\gg10^{-2}$ karyotypes will acquire lethal nullisomies)}. \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. As a result, the fraction of cancer cells with aneuploidy \textcolor{green}{of interest} is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy \textcolor{green}{of interest}).

We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy \textcolor{green}{other than} trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our \textcolor{green}{analysis}, we assume that lineages produced by cells from the initial population divide and die independently of each other.
\textcolor{green}{However}, the cells will compete for resources, but we expect that this \textcolor{green}{competition} can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which \textcolor{green}{competition is most} important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity \textcolor{green}{in} this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the \textcolor{green}{tumor} population \textcolor{green}{size}, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second case, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before \textcolor{green}{the} onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore \textcolor{green}{we} use $\lambda_w$ for the wildtype division rate in the drug-free environment).

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if wildtype cells \textcolor{green}{die rapidly} (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.  
In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs. \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$. Therefore, standing genetic variation does not drive evolution of drug resistance when compared to \emph{de-novo} aneuploidy, but it does offer a slight advantage when compared with direct mutation.

@@

When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long. We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the dynamics of \textcolor{green}{the} number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size. % TODO I dont like the word "behaves", please be specific. - what exactly is deterministic and why does it mean you cannot do the separation. ok
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

@@

Additionally, our model predictions can be tested with data from patients. Given the type of cancer and how \textcolor{green}{beneficial} aneuploidy is,  our model can predict the probability for cancer relapse after treatment.

>>>>

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: Yoav Ram (e-mail: yoavram@tauex.tau.ac.il)}

@@

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments with yeast and bacteria.

Here we study evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone towards resistant cancer cells. We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

The cell population at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):

@@

We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division. 
\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division (for $u\ll10^{-3}$ no adaptation occurs and for $u\gg10^{-2}$ cells will acquire lethal nullisomies where both pairs of chromosomes are lost). \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. As a result, the fraction of cancer cells with aneuploidy of interest is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy of interest).

We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analysis, we assume that lineages produced by cells from the initial population divide and die independently of each other.
However, the cells will compete for resources, but we expect that this competition can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which competition is most important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following scenarios, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease of the tumor population size, allowing more opportunities for producing resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
In the first scenario, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second scenario. In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.

We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore we use $\lambda_w$ for the wildtype division rate in the drug-free environment).

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if wildtype cells die rapidly (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.  
In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs. \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid scenario and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$. Therefore, standing genetic variation does not drive evolution of drug resistance when compared to \emph{de-novo} aneuploidy, but it does offer a slight advantage when compared with direct mutation.

@@

When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long. We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
However, when the expected number of resistant lineages is large, the dynamics of the number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size. % TODO I dont like the word "behaves", please be specific. - what exactly is deterministic and why does it mean you cannot do the separation. ok
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

@@

Additionally, our model predictions can be tested with data from patients. Given the type of cancer and how beneficial aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
